Approximation as Differentiation

### 2025-07-08 main/introduction.tex
vákár -> vakar (easier).

Previous work on \GPS used a special operational semantics to generate a trace of each execution, and then separately used that trace to compute the Galois connection described above, by re-running forwards or backwards over the trace. It would be useful to have a denotational account of \GPS as well, especially if we could provide a semantics where the backwards analysis is baked in, rather than provided by a separately (but somewhat redundantly) defined ``backwards evaluation'' operation. As it happens there is a close analogy between \GPS and \emph{automatic differentiation} for differentiable programs~\cite{siskind08,elliott18,vákár22}, which points to a way to develop such an approach. We have already hinted at this in the description above, but let us now make it explicit.

>>>>

Previous work on \GPS used a special operational semantics to generate a trace of each execution, and then separately used that trace to compute the Galois connection described above, by re-running forwards or backwards over the trace. It would be useful to have a denotational account of \GPS as well, especially if we could provide a semantics where the backwards analysis is baked in, rather than provided by a separately (but somewhat redundantly) defined ``backwards evaluation'' operation. As it happens there is a close analogy between \GPS and \emph{automatic differentiation} for differentiable programs~\cite{siskind08,elliott18,vakar22}, which points to a way to develop such an approach. We have already hinted at this in the description above, but let us now make it explicit.

### 2025-07-08 main/models.tex
vákár -> vakar (easier).

As similar result is given by \citet{cruttwell2022}, where Euclidean
spaces $\RR^n$ and smooth functions are embedded into a category of
lenses (the ``simply typed'' version of the $\Fam$ construction). As
in \citet{vákár22}, the idea is to formally separate functions on
points and their forward/reverse tangent maps for the purposes of
implementation of automatic differentiation. In the case of smooth
maps, this process throws away information on higher derivatives by
turning smooth maps into pairs of plain functions and linear
functions. We conjecture at the end of this section that the analogous
construction in our partially ordered setting does not.

>>>>

As similar result is given by \citet{cruttwell2022}, where Euclidean
spaces $\RR^n$ and smooth functions are embedded into a category of
lenses (the ``simply typed'' version of the $\Fam$ construction). As
in \citet{vakar22}, the idea is to formally separate functions on
points and their forward/reverse tangent maps for the purposes of
implementation of automatic differentiation. In the case of smooth
maps, this process throws away information on higher derivatives by
turning smooth maps into pairs of plain functions and linear
functions. We conjecture at the end of this section that the analogous
construction in our partially ordered setting does not.

### 2025-07-08 main/related-work.tex
vákár -> vakar (easier).

More recent approaches to automatic differentiation have emphasised semantic foundations. \citet{elliott18}
proposed a categorical model of AD that interprets programs as functions enriched with their derivatives,
giving a compositional account of differentiation based on duality and linear maps. Vákár and
collaborators~\cite{vákár22,nunes2023} developed the CHAD framework which inspired this paper, using
Grothendieck constructions over indexed categories to capture both values and their tangents in a
compositional semantic structure. These perspectives shed light on the categorical structure of AD and guide
the design of systems that generalise AD beyond real analysis, including the application to data provenance
and slicing explored in this paper.

>>>>

More recent approaches to automatic differentiation have emphasised semantic foundations. \citet{elliott18}
proposed a categorical model of AD that interprets programs as functions enriched with their derivatives,
giving a compositional account of differentiation based on duality and linear maps. Vákár and
collaborators~\cite{vakar22,nunes2023} developed the CHAD framework which inspired this paper, using
Grothendieck constructions over indexed categories to capture both values and their tangents in a
compositional semantic structure. These perspectives shed light on the categorical structure of AD and guide
the design of systems that generalise AD beyond real analysis, including the application to data provenance
and slicing explored in this paper.

### 2025-07-08 old/main/first-order.tex
Purging section 4.
vákár -> vakar (easier).

Our denotational approach to \GPS is based on the CHAD (Combinatory Homomorphic Automatic Differentiation) work of Mattias Vákár and others~\cite{vákár22,nunes2023}. Before we go into details, we present a high-level overview of their approach and how we have applied it to \GPS.

@@

We now consider how to adapt the CHAD framework~\cite{vákár22,nunes2023} to interpret programs as functions
along with their forward and backward approximation maps. Since the appropriate lattice of approximation
typically depends on the value being approximated, we will want to interpret a type $\tau$ as a pair $(X,
\partial X)$ where $X$ provides the conventional interpretation as a set and $\partial X$ provides an
appropriate lattice of approximations $\partial X(x)$ for every point $x \in X$. An expression $e$ of type
$\sigma$ with a single free variable of type $\tau$ (supposing $\tau$ and $\sigma$ denote $X$ and $Y$ in the
conventional semantics) must then be interpreted by a pair $(f, \partial f)$, with $f$ providing the
conventional interpretation as a function $X \to Y$, and $\partial f$ as a family of Galois connections
$\partial f(x): \partial X(x) \to \partial Y(f(x))$ taking approximations of $x$ to approximations of $f(x)$.

@@

% \subsubsection{Derivatives of Differentiable Functions}
%
% For flat manifolds like $\RR^n$, the tangent space and cotangent space at a point $x$ are both canonically
% isomorphic to $\RR^n$, so the forward and backward derivatives of a differentiable function $f: \RR^m \to
% \RR^n$ at a point $x$ have the following form \todo{Can probably delete now as covered in
% \secref{approx-as-tangents}}:
%
% \begin{itemize}
% \item The \emph{forward derivative} (tangent map or pushforward) $\pushf{f}_x$ of $f$ at $x \in \RR^m$ is the
% unique linear map $\RR^m \linearto \RR^n$ given by the Jacobean matrix of $f$ at $x$.
% \item The \emph{backward derivative} (cotangent map or pullback) $\pullf{f}_x$ is the unique linear map
% $\RR^n \linearto \RR^m$ given by the transpose (adjoint) of the Jacobean matrix of $f$ at $x$.
% \end{itemize}

>>>>

Our denotational approach to \GPS is based on the CHAD (Combinatory Homomorphic Automatic Differentiation) work of Mattias Vákár and others~\cite{vakar22,nunes2023}. Before we go into details, we present a high-level overview of their approach and how we have applied it to \GPS.

@@

We now consider how to adapt the CHAD framework~\cite{vakar22,nunes2023} to interpret programs as functions
along with their forward and backward approximation maps. Since the appropriate lattice of approximation
typically depends on the value being approximated, we will want to interpret a type $\tau$ as a pair $(X,
\partial X)$ where $X$ provides the conventional interpretation as a set and $\partial X$ provides an
appropriate lattice of approximations $\partial X(x)$ for every point $x \in X$. An expression $e$ of type
$\sigma$ with a single free variable of type $\tau$ (supposing $\tau$ and $\sigma$ denote $X$ and $Y$ in the
conventional semantics) must then be interpreted by a pair $(f, \partial f)$, with $f$ providing the
conventional interpretation as a function $X \to Y$, and $\partial f$ as a family of Galois connections
$\partial f(x): \partial X(x) \to \partial Y(f(x))$ taking approximations of $x$ to approximations of $f(x)$.

@@

% \subsubsection{Derivatives of Differentiable Functions}
%
% For flat manifolds like $\RR^n$, the tangent space and cotangent space at a point $x$ are both canonically
% isomorphic to $\RR^n$, so the forward and backward derivatives of a differentiable function $f: \RR^m \to
% \RR^n$ at a point $x$ have the following form \roly{Can probably delete now as covered in
% \secref{approx-as-tangents}}:
%
% \begin{itemize}
% \item The \emph{forward derivative} (tangent map or pushforward) $\pushf{f}_x$ of $f$ at $x \in \RR^m$ is the
% unique linear map $\RR^m \linearto \RR^n$ given by the Jacobean matrix of $f$ at $x$.
% \item The \emph{backward derivative} (cotangent map or pullback) $\pullf{f}_x$ is the unique linear map
% $\RR^n \linearto \RR^m$ given by the transpose (adjoint) of the Jacobean matrix of $f$ at $x$.
% \end{itemize}
